### 2017-06-29 motivation.tex
Edits.

Note that, in the above example, there is no Write-Write conflict, or
the \emph{lost update} anomaly that often characterizes concurrency
bugs, such as illegal concurrent withdraws on a bank
account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
\C{d\_next\_o\_id} field of the district record, they do so atomically
(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
present in the final state. Likewise, four other well-known anomalies
that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
are also not exhibited by the example. Thus, any program analysis,
static or dynamic, that aims to determine appropriate isolation by
checking for possible manifestations of known anomalies fails to
identify grounds for promoting the isolation level of \C{new\_order}.
Yet, if we take the semantics of the application into account, it is
clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, there is a need for a high-level verification
framework that lets us reason about transaction semantics under weak
isolation.

>>>>

Note that, in the above example, there is no Write-Write conflict, or
the \emph{lost update} anomaly that often characterizes concurrency
bugs, such as illegal concurrent withdraws on a bank
account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
\C{d\_next\_o\_id} field of the district record, they do so atomically
(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
present in the final state. Likewise, four other well-known anomalies
that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
are also not exhibited by the example. Thus, any program analysis,
static or dynamic, that aims to determine appropriate isolation by
checking for possible manifestations of known anomalies fails to
identify grounds for promoting the isolation level of \C{new\_order}.
Yet, if we take the semantics of the application into account, it is
clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, there is a need for a high-level verification
framework that lets us reason about transaction semantics in
conjunction with the semantics of weak isolation.

Facilitating high-level reasoning about weak isolation is however not
easy owing to the low-level nature of its specification, either in
terms of the implementation details~\cite{gray}, allowed
anomalies~\cite{berenson}, or well-formedness condition on program
traces~\cite{adyaphd,gotsmanconcur15}. For instance, does promoting
the isolation level of \C{new\_order} to Repeatable Read (RR) help in
preserving the consistency condition? To answer this question, one has
to relate the low-level specification of RR given in terms of
trace-based relations~\cite{adyaphd,gotsmanconcur15} to the semantics
of TPC-C, which is difficult. Alternatively, one could reason about
how RR is implemented, say on Postgres, and reason about the
\C{new\_order} implementation in conjunction with Postgres's RR
implementation. However, this is not any easier, considering that
Postgres employs a host of techniques, including Multiversion, locks
and rollbacks, to implement RR. The verification problem becomes even
more difficult if the aim is to facilitate automation. Our experience
suggests that a proof framework that requires programmers to relate
trace-based isolation specifications to program semantics necessarily
imposes heavy annotation burden, requires painstakingly long and
verbose proofs, and does not lend itself to automation. Weak isolation
aside, there are also the conventional challenges that face
(automatic) verification of concurrent programs: facilitating modular
proofs, inferring intermediary assertions and loop invariants that are
stable w.r.t interference, etc. A proof system for weakly isolated
transactions has to necessarily address these challenges, along with
those that arise due to weak isolation.

The proof system and the automation framework proposed in this paper
overcomes or circumvents the aforementioned problems. The first
observation that underlies our approach is that, while databases
implement weak isolation employing a host of techniques to facilitate
maximum concurrency despite the isolation, the operational semantics
of such implementations can nonetheless be captured using a simple
abstraction of the database state exposed to the application. For
instance, consider how Postgres executes an RR transaction. A
(conceptual) snapshot of the database state is created at the
beginning of the transaction against which its reads are executed.
When the transaction attempts to update a record, Postgres first
checks if the current version of the record in the database is same as
the snapshot version. If yes, then an exclusive lock is obtained on
the record, and the updatation is carried out. If the record is
already locked, Postgres waits (blocks the transaction) until the lock
is released. Once obtained, an exclusive lock is held until the
transaction commits or rolls back. If the aforementioned version check
fails (i.e., database has a later version than the snapshot), then the
transaction is rolled back and re-executed. Thus, the implementation
of RR on Postgres is fairly involved. Nonetheless, its operational
semantics can be described straightforwardly in terms of the legal
transitions of the database state (denoted, $\Delta \longrightarrow
\Delta'$) allowed while an RR transaction is in progress. Since
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state doesn't
change (i.e., $\Delta' = \Delta$) during the execution. Note that the
uncommitted writes of the transaction are not part of $\Delta$;
instead we imagine a transaction-local state ($\delta$) that is
populated with only the uncommitted local writes. When the transaction
commits, the local state ($\delta$) is flushed to the global state
($\Delta$) to yeild a committed state. However, unlike reads, commit
is performed against the current state of the database (call it
$\Delta_c$); not the snapshot. Thus, after the transaction finishes
execution, but before it commits, the transaction witnesses the
effects of all concurrent transactions in a single database state
transition $\Delta \longrightarrow \Delta_c$. Postgres's RR
implementation effectively constrains this transition. Firstly, due to
the version check, the current transaction cannot update a data item
that was already updated by a concurrent transaction. Secondly, due to
exclusive write locks, a data item updated by the current transaction
cannot be overwritten by a concurrent transaction. Putting these two
together, we know that the transition $\Delta \longrightarrow
\Delta_c$ does not involve any concurrent transactions that write to
the same data items as the current RR transaction. Since $\delta$
records the writes of the current transaction, intuitively, forall
$x\in\delta$, $\Delta_c(x) = \Delta(x)$. To summarize, the operational
semantics of Postgres's RR implementation can be captured as
constraints on the transitions of the database state ($\Delta
\longrightarrow \Delta'$) during the lifetime an RR
transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}
